The error in Bayesian phylogenetic reconstruction when speciation is not instantaneous

### 2018-05-14 article.tex
Feedback

Current phylogenetic tools use tree priors 
that assume speciation is instantaneous, whilst we know that, 
speciation is often a gradual process \notebf{[REF]}.
The (constant-rate) birth-death (BD) model is a commonly 
used tree prior, but it ignores this temporal aspect of speciation.
The protracted birth-death (PBD) model, an extension of 
the BD model, does incorporate the idea that speciation takes time.
In this model, a branching event does not give rise to a new species, but to
a new species-to-be, called an incipient species. Such an incipient
species may go extinct, finish its speciation to become a good species, or give
rise to new incipient species. Protracted speciation may explain observed 
declines in lineage accumulation (\cite{etienne_and_rosindell_2012}).

@@

The PBD model has five biological parameters (see \ref{table:parameters}), 
which we explore in a factorial fashion, excluding some combinations. 
We only simulate a PBD process for phylogenies in which
speciation initiation exceeds extinction rate ($b_i > \mu_i$ and $b_g > \mu_g$),
and in which the expected number of extant good species is less than 1000.
{NOTE: accoring to Rampal this has been solved analytically (for the Protracted Birth and Death model). Where?}.
We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
The parameter values chosen are based on the parameter sets used by 
\cite{etienne2014}, as these parameters were shown to result in reasonably
sized phylogenies and using the same set allows us to compare results. 
For the speciation initiation rates of good and incipient species, $b_g$ and $b_i$ 
respectively, we use $0.1$, $0.5$ and $1.0$ speciation initiation events per good/incipient species per time unit.
The speciation completion rates we use are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
events per (incipient species) species per time unit. We use $10^9 \approx \infty$
to mimic the BD model, because the PBD model reduces to the BD model for $\lambda = \infty$.
This allows us to measure the baseline error, 
which is the difference between inferred tree and true species tree that arises purely due to noise 
because the generating model and the model used in inference are identical in this case.
The extinction rates of good and incipient species, $\mu_g$ and $\mu_i$ respectively, 
that we use are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per good/incipient species per time unit.

From each biological parameter set, we simulate a protracted birth-death tree,
using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
all with a crown age of 15 million years.
Each protracted birth-death tree uses a different random number
generatior seed, which makes all runs independent, resulting in a balanced 
data set. \textbf{NOTE: maybe discuss with Rampal to remove confusion. I see if I can get the
students to make a histogram of number of taxa in the incipient species trees, as Rampal
requires a geometric distribution}

@@

From an alignment, we run a Bayesian analysis and 
create a posterior distribution of trees and parameters
using the \verb;babette; (\cite{babette}) package
that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
For our site model, we assume a Jukes-Cantor 
nucleotide substitution model, as used in the simulation of the alignment.
For our clock model, we assume a strict clock with the same fixed rate as 
used in the simulation of the alignment. 
The tree prior assumed in inference is the BD model, 
because studying the effect of this assumption is the goal of this study. 
We assume an MRCA prior with a tight normal distribution
around the crown age, by choosing the crown age as mean, and a standard deviation 
of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
units) as the alignment. 
We ran the MCMC chain to generate 1111 states,
of which we remove the first 10\% (also called the 'burn-in'). 
Of the remaining
1000 MCMC states, the effective sample size (ESS) of the posterior \textbf{NOTE: there
is a parameter estimate called 'posterior'. I choose to pick that one, and I assume
it is the wiser choice over 'prior' and parameter estimates. Must discuss} must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
If the ESS is less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.

>>>>

Current phylogenetic tools use tree priors 
that assume speciation is instantaneous, whilst we know that, 
speciation is often a gradual process (\cite{schluter2009}).
The (constant-rate) birth-death (BD) model is a commonly 
used tree prior, but it ignores this temporal aspect of speciation.
The protracted birth-death (PBD) model, an extension of 
the BD model, does incorporate the idea that speciation takes time.
In this model, a branching event does not give rise to a new species, but to
a new species-to-be, called an incipient species. Such an incipient
species may go extinct, finish its speciation to become a good species, or give
rise to new incipient species. Protracted speciation may explain observed 
declines in lineage accumulation (\cite{etienne_and_rosindell_2012}).

@@

The PBD model has five biological parameters (see \ref{table:parameters}), 
which we explore in a factorial fashion, excluding some combinations. 
We only simulate a PBD process for phylogenies in which
speciation initiation exceeds extinction rate ($b_i > \mu_i$ and $b_g > \mu_g$),
and in which the expected number of extant good species is less than 1000.
\textbf{[NOTE: accoring to Rampal this has been solved analytically (for the Protracted Birth and Death model). Where?]}.
We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
The parameter values chosen are based on the parameter sets used by 
\cite{etienne2014}, as these parameters were shown to result in reasonably
sized phylogenies and using the same set allows us to compare results. 
For the speciation initiation rates of good and incipient species, $b_g$ and $b_i$ 
respectively, we use $0.1$, $0.5$ and $1.0$ speciation initiation events per good/incipient species per time unit.
The speciation completion rates we use are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
events per (incipient species) species per time unit. We use $10^9 \approx \infty$
to mimic the BD model, because the PBD model reduces to the BD model for $\lambda = \infty$.
This allows us to measure the baseline error, 
which is the difference between inferred tree and true species tree that arises purely due to noise 
because the generating model and the model used in inference are identical in this case.
The extinction rates of good and incipient species, $\mu_g$ and $\mu_i$ respectively, 
that we use are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per good/incipient species per time unit.

From each biological parameter set, we simulate a protracted birth-death tree,
using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
all with a crown age of 15 million years.
Each protracted birth-death tree uses a different random number
generatior seed, which makes all runs independent, resulting in a balanced 
data set. \textbf{[NOTE: maybe discuss with Rampal to remove confusion. Rampal requests a histogram of branch lengths to follow a geometric distribution]}

@@

From an alignment, we run a Bayesian analysis and 
create a posterior distribution of trees and parameters
using the \verb;babette; (\cite{babette}) package
that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
For our site model, we assume a Jukes-Cantor 
nucleotide substitution model, as used in the simulation of the alignment.
For our clock model, we assume a strict clock with the same fixed rate as 
used in the simulation of the alignment. 
The tree prior assumed in inference is the BD model, 
because studying the effect of this assumption is the goal of this study. 
We assume an MRCA prior with a tight normal distribution
around the crown age, by choosing the crown age as mean, and a standard deviation 
of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
units) as the alignment. 
We ran the MCMC chain to generate 1111 states,
of which we remove the first 10\% (also called the 'burn-in'). 
Of the remaining
1000 MCMC states, the effective sample size (ESS) of the posterior \textbf{[NOTE: there
is a parameter estimate called 'posterior'. I choose to pick that one, and I assume
it is the wiser choice over 'prior' and parameter estimates. Must discuss]} must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
If the ESS is less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.
